\chapter{The multiplicative monoid \texorpdfstring{$\Nplus$}{N+}}

The Arithmetic Site, introduced by Connes and Consani, is a geometric
object — a topos equipped with a structure sheaf — whose points in the
topos-theoretic sense are in natural bijection with the adèle class space
of $\mathbb{Q}$.  Its construction begins with a single, elementary
monoid: the positive integers under multiplication.  Everything else —
the tropical semiring, the presheaf topos, the structure sheaf, and
ultimately the connection to the Riemann zeta function — is built on top
of this one piece of data.

\begin{definition}[label=def:Nplus]
  \label{def:Nplus}\lean{ArithmeticSite.NPos}\leanok
  Let $\Nplus$ denote the set $\{1, 2, 3, \ldots\}$ of strictly positive
  integers equipped with ordinary multiplication. This is a commutative
  monoid with unit $1$.
  In Lean we represent this as \texttt{PNat} with its multiplicative
  structure, via the abbreviation \texttt{NPos}.
\end{definition}

The unique factorisation theorem says that $\Nplus$ is the free
commutative monoid on the set of primes: every element factors uniquely
as a finite product $\prod_p p^{a_p}$.  This arithmetic richness ---
the interplay of primes, divisibility, and factorisation --- is what
makes $\Nplus$ the right monoid for the Arithmetic Site.  As we will
see, it is ultimately responsible for the fact that the points of
$\PSh{\Nplus}$ are parametrised by the ad\`ele class space of
$\mathbb{Q}$, and that the Hasse--Weil zeta function of the site
recovers the Riemann zeta function.

\begin{definition}[label=def:BNplus, uses=def:Nplus]
  \label{def:BNplus}\uses{def:Nplus}\lean{ArithmeticSite.BNplus}\leanok
  The \emph{classifying category} $B\Nplus$ of $\Nplus$ is the
  one-object category whose single object is $*$ and whose morphisms
  $* \to *$ are the elements of $\Nplus$, with composition given by
  multiplication.
\end{definition}

\begin{remark}
  The passage from a monoid $M$ to the one-object category $BM$ is the
  standard \emph{delooping} construction.  A presheaf on $BM$ — that is,
  a functor $BM^{\mathrm{op}} \to \Cat{Set}$ — is exactly a set $X$
  equipped with a left action of $M$: each element $n \in M$ acts as a
  function $X \to X$, and the action is compatible with composition.
  (Left and right actions coincide here since $\Nplus$ is commutative.)
  The presheaf topos $\PSh{\Nplus}$ introduced in
  Chapter~\ref{chap:presheaf_topos} is precisely the category of such
  sets-with-$\Nplus$-action, and it is in this topos that the Arithmetic
  Site lives.
\end{remark}

\chapter{The tropical semiring \texorpdfstring{$\Nbar$}{Nbar}}

The tropical semiring $\Nbar$ is the second fundamental ingredient of
the Arithmetic Site.  It plays the role of the \emph{structure sheaf}:
as Connes and Consani put it, $\Nbar$ is the semiring on which $\Nplus$
acts by multiplication, and it is this action that makes $\Nbar$ into
a sheaf of semirings over the topos $\widehat{B\Nplus}$.  The arithmetic
of $\Nbar$ is characteristic-$1$ arithmetic --- addition is idempotent
--- which is the tropical, or min-plus, shadow of ordinary arithmetic.
Connes and Consani define $\Nbar$ explicitly as $(\mathbb{N} \cup \{\infty\},
\inf, +)$, confirming the min-plus convention: addition is $\min$, with
$+\infty$ as the additive identity.  When elements of $\Nbar$ are written
as exponents $q^x$ for $q \in (0,1)$, the addition becomes $\max$ on
values (since smaller exponent means larger value), which is why the
stalks in the paper appear in max-plus form as $[0,1]_{\max}$ and
$\mathbb{R}_{\max}$.

\begin{definition}[label=def:Nbar]
  \label{def:Nbar}\lean{ArithmeticSite.NBar}\leanok
  The \emph{tropical semiring} $\Nbar$ is the set
  $\mathbb{N} \cup \{+\infty\}$ equipped with:
  \begin{itemize}
    \item \emph{Tropical addition}: $a \oplus b := \min(a, b)$, with
          identity element $+\infty$.
    \item \emph{Tropical multiplication}: $a \odot b := a + b$
          (ordinary addition), with identity element $0$.
  \end{itemize}
  In Lean, $\Nbar$ is represented as \texttt{Tropical (WithTop $\mathbb{N}$)},
  via the abbreviation \texttt{NBar}.
  The zero element is \texttt{trop $\top$} (i.e.\ $+\infty$) and the
  multiplicative identity is \texttt{trop 0}.
\end{definition}

\begin{lemma}[label=lem:Nbar_semiring, uses=def:Nbar]
  \label{lem:Nbar_semiring}\uses{def:Nbar}\lean{ArithmeticSite.NBar_commSemiring}\leanok
  $\Nbar$ is a commutative semiring.
\end{lemma}
\begin{proof}
  This is the Mathlib instance
  \texttt{CommSemiring (Tropical (WithTop $\mathbb{N}$))},
  found by \texttt{inferInstance}.
\end{proof}

\begin{lemma}[label=lem:Nbar_idempotent, uses=def:Nbar]
  \label{lem:Nbar_idempotent}\uses{def:Nbar}\lean{ArithmeticSite.NBar_add_idempotent}\leanok
  Tropical addition on $\Nbar$ is idempotent: $a \oplus a = a$ for all
  $a \in \Nbar$.
\end{lemma}
\begin{proof}
  $a \oplus a = \min(a,a) = a$. In Lean this follows from
  \texttt{Tropical.add\_eq\_left}.
\end{proof}

\begin{remark}
  Idempotency of addition --- equivalently, the identity $1 \oplus 1 = 1$
  --- is the defining feature of \emph{characteristic $1$} arithmetic.
  It is what places $\Nbar$ in the same conceptual family as finite fields
  $\mathbb{F}_p$ (characteristic $p$), with $\min$ playing the role that
  $+\bmod p$ plays there.  The scaling endomorphisms defined below are the
  characteristic-$1$ analogue of the Frobenius endomorphism $x \mapsto x^p$
  of $\mathbb{F}_p$.
\end{remark}

\begin{definition}[label=def:scaling_action, uses=def:Nbar, def:Nplus]
  \label{def:scaling_action}\uses{def:Nbar, def:Nplus}\lean{ArithmeticSite.scalingEndomorphism}\leanok
  For each $n \in \Nplus$, the \emph{scaling endomorphism}
  $\varphi_n : \Nbar \to \Nbar$ is defined by
  \[
    \varphi_n(x) := nx \quad (x \in \mathbb{N}),
    \qquad \varphi_n(+\infty) := +\infty,
  \]
  where $nx$ denotes ordinary multiplication in $\mathbb{N}$.
  Equivalently, $\varphi_n(x) = x^{\odot n}$, the $n$-fold tropical
  product of $x$ with itself.
  In Lean, $\varphi_n$ is defined as a semiring endomorphism
  \texttt{NBar~->+*~NBar} via \texttt{scalingEndomorphism}.
\end{definition}

\begin{lemma}[label=lem:scaling_semiring_hom, uses=def:scaling_action, lem:Nbar_semiring]
  \label{lem:scaling_semiring_hom}\uses{def:scaling_action, lem:Nbar_semiring}\lean{ArithmeticSite.scalingEndomorphism}\leanok
  For each $n \in \Nplus$, the map $\varphi_n$ is a semiring endomorphism
  of $\Nbar$: it preserves both operations and both identities.
\end{lemma}
\begin{proof}
  Preservation of $+\infty$ and of tropical multiplication are immediate.
  Preservation of tropical addition uses $n \cdot \min(a,b) = \min(na,nb)$
  for $n > 0$, which holds because scaling by $n$ is order-preserving on
  $\mathbb{N} \cup \{+\infty\}$.  In Lean this is the \texttt{map\_add'}
  field of \texttt{scalingEndomorphism}, proved via monotonicity of
  \texttt{nsmul} and a case split on \texttt{le\_total}.
\end{proof}

\begin{lemma}[label=lem:scaling_comp, uses=def:scaling_action]
  \label{lem:scaling_comp}\uses{def:scaling_action}\lean{ArithmeticSite.scalingEndomorphism_comp}\leanok
  The scaling endomorphisms compose multiplicatively:
  \[
    \varphi_m \circ \varphi_n = \varphi_{mn}
  \]
  for all $m, n \in \Nplus$.  In particular, the map $n \mapsto \varphi_n$
  is a monoid homomorphism from $(\Nplus, \times)$ into the semiring
  endomorphism monoid $\mathrm{End}_{\mathrm{Semiring}}(\Nbar)$.
\end{lemma}
\begin{proof}
  $(\varphi_m \circ \varphi_n)(x) = m(nx) = (mn)x = \varphi_{mn}(x)$.
  In Lean this is \texttt{scalingEndomorphism\_comp}, proved via
  \texttt{pow\_mul}.
\end{proof}

\begin{remark}
  The Frobenius \emph{correspondences} $\Psi(\lambda)$, parametrised by
  $\lambda \in \Qplus$, are a more refined object: they live in the
  \emph{square} $\widehat{B{\Nplus}^2}$ of the Arithmetic Site and are
  defined via the action of $\Nplus \times \Nplus$ on
  $\Nbar \otimes_{\mathbb{B}} \Nbar$.  They are introduced in
  Chapter~\ref{chap:frobenius_correspondences}.
\end{remark}

\chapter{The presheaf topos \texorpdfstring{$\PSh{\Nplus}$}{PSh(N+)}}
\label{chap:presheaf_topos}

The topos underlying the Arithmetic Site is the simplest kind: a
presheaf topos.  Its objects are sets equipped with a left
$\Nplus$-action by arbitrary functions --- not necessarily bijections,
since $\Nplus$ is a monoid rather than a group --- and its morphisms are
$\Nplus$-equivariant maps between them.  The multiplicative structure of
$\Nplus$ is entirely encoded in the composition of these actions: if $n$
and $m$ act on a set $X$, then $nm$ acts as the composite $x \mapsto
n \cdot (m \cdot x)$.

Connes and Consani note that $\PSh{\Nplus} \simeq
\mathrm{Sh}(\Nplus, J)$, where $J$ is the \emph{chaotic topology} on
$\Nplus$ --- the topology in which every presheaf is already a sheaf.
This means there are no genuine gluing conditions to check: the topos is
simply the presheaf category, with no further data.  The covering
families arising from factorizations of integers, foreshadowed in
Chapter~1, reflect a finer site structure that one \emph{can} put on
$\Nplus$; the Arithmetic Site itself uses the coarser chaotic topology,
where that combinatorial structure is not needed.

\begin{definition}[label=def:presheaf_topos, uses=def:BNplus]
  \label{def:presheaf_topos}\uses{def:BNplus}\lean{ArithmeticSite.PresheafTopos}\leanok
  The \emph{presheaf topos} $\PSh{\Nplus}$ is the category of functors
  $B{\Nplus}^{\mathrm{op}} \to \Cat{Set}$, i.e.\ the category of sets
  equipped with a left $\Nplus$-action. In Lean this is
  \texttt{Action (Type*) (Multiplicative $\mathbb{N}$+)}.
\end{definition}

\begin{lemma}[label=lem:presheaf_topos_is_topos, uses=def:presheaf_topos]
  \label{lem:presheaf_topos_is_topos}\uses{def:presheaf_topos}\lean{ArithmeticSite.presheafTopos_isGrothendieckTopos}
  $\PSh{\Nplus}$ is a Grothendieck topos.
\end{lemma}
\begin{proof}
  Any presheaf category on a small category is a Grothendieck topos.
  The site is $(\Nplus, J)$ where $J$ is the chaotic topology.
\end{proof}

\begin{remark}
  The points of $\PSh{\Nplus}$ --- geometric morphisms from $\Cat{Set}$
  to $\PSh{\Nplus}$ --- are the deep link to number theory.  Connes and
  Consani prove (Theorem~2.2 of their 2014 paper) that the category of points
  of $\PSh{\Nplus}$ is canonically equivalent to the category of totally
  ordered groups isomorphic to non-trivial subgroups of
  $(\mathbb{Q}, \mathbb{Q}_+)$.  The space of isomorphism classes of
  such points is canonically isomorphic to the double quotient
  \[
    \Qplus \backslash \Af / \Zhat^*,
  \]
  where $\Af$ is the ring of finite ad\`eles of $\mathbb{Q}$ and
  $\Zhat^* = \hat{\mathbb{Z}}^\times$ is the group of units of the
  profinite completion of $\mathbb{Z}$.  This quotient is a piece of the
  ad\`ele class space of $\mathbb{Q}$, the same space that appears in
  Connes' noncommutative-geometric approach to the Riemann
  hypothesis.  The structure sheaf $\mathcal{O} = \Nbar$ then equips
  this space with characteristic-$1$ geometry, whose Hasse--Weil zeta
  function recovers the Riemann zeta function.
\end{remark}

\chapter{The structure sheaf}

In classical algebraic geometry, a \emph{ringed space} is a topological
space equipped with a sheaf of commutative rings --- the \emph{structure
sheaf} --- whose stalks encode the local algebra of the space.  The
Arithmetic Site is a \emph{ringed topos} in the same spirit: the role of
the topological space is played by the presheaf topos $\PSh{\Nplus}$,
and the role of the structure sheaf is played by $\Nbar$, viewed as a
semiring object living inside that topos.

An object of $\PSh{\Nplus}$ is a set equipped with a left
$\Nplus$-action, so for $\Nbar$ to be a semiring \emph{object} of
$\PSh{\Nplus}$ the action must be by semiring endomorphisms: each
scaling map $\varphi_n$ must preserve both tropical operations and both
identities.  This is exactly the content of
Lemma~\ref{lem:scaling_semiring_hom}, which is therefore the key
structural fact underlying the definition below.

\begin{definition}[label=def:structure_sheaf,
                  uses=def:presheaf_topos, def:Nbar, def:scaling_action]
  \label{def:structure_sheaf}\uses{def:presheaf_topos, def:Nbar, def:scaling_action}\lean{ArithmeticSite.structureSheaf}
  The \emph{structure sheaf} $\mathcal{O}$ of the Arithmetic Site is the
  object of $\PSh{\Nplus}$ whose underlying set is $\Nbar$ and whose
  $\Nplus$-action is given by the scaling endomorphisms:
  $n \cdot x := \varphi_n(x) = nx$.
\end{definition}

\begin{lemma}[label=lem:structure_sheaf_action,
             uses=def:structure_sheaf, lem:scaling_comp, lem:scaling_semiring_hom]
  \label{lem:structure_sheaf_action}\uses{def:structure_sheaf, lem:scaling_comp, lem:scaling_semiring_hom}\lean{ArithmeticSite.structureSheaf_isMonoidHom}\leanok
  The assignment $n \mapsto \varphi_n$ defines a monoid homomorphism
  $(\Nplus, \times) \to \mathrm{End}_{\mathrm{Semiring}}(\Nbar)$,
  making $\Nbar$ into a semiring object of $\PSh{\Nplus}$ with
  $\Nplus$ acting by multiplication.
\end{lemma}
\begin{proof}
  Each $\varphi_n$ is a semiring endomorphism by
  Lemma~\ref{lem:scaling_semiring_hom}.  The homomorphism property
  $\varphi_{mn} = \varphi_m \circ \varphi_n$ is
  Lemma~\ref{lem:scaling_comp}, and $\varphi_1 = \mathrm{id}$ since
  $\varphi_1(x) = 1 \cdot x = x$.  In Lean this is
  \texttt{scalingEndomorphism\_comp}.
\end{proof}

\begin{remark}
  Since the Arithmetic Site uses the chaotic topology, every presheaf on
  $B\Nplus$ is automatically a sheaf --- there are no gluing conditions
  to verify.  The content of the structure sheaf is therefore entirely
  algebraic: $\Nbar$ is not merely a set with $\Nplus$-action, but a
  semiring whose operations are \emph{preserved} by that action.  This
  is what Lemma~\ref{lem:structure_sheaf_action} records.
\end{remark}

\begin{remark}
  Connes and Consani (Theorem~2.4 of their 2014 paper) describe the
  stalks of $\mathcal{O}$ explicitly.  At the point corresponding to an
  intermediate semifield $F \subset K \subset \bar{F}$, the stalk is
  \[
    \mathcal{O}_K = \{ r \in K \mid r \oplus 1 = 1 \},
  \]
  the subsemiring of elements that are ``bounded above by $1$'' in the
  ordering of $K$.  This is the characteristic-$1$ analogue of the ring of integers of a
  local field: just as the ring of integers of a $p$-adic field $K_v$
  consists of elements of norm at most $1$, the stalk
  of $\mathcal{O}$ picks out the elements ``bounded above by $1$''.
  At the ``generic'' point (corresponding to $K = F = Z_{\max}$) the
  stalk recovers $\Nbar$ itself.
\end{remark}

\chapter{The Arithmetic Site}

All the ingredients are now in place.  The presheaf topos $\PSh{\Nplus}$
provides the geometric stage: a Grothendieck topos whose points are
parametrised by the ad\`ele class space of $\mathbb{Q}$.  The tropical
semiring $\Nbar$, equipped with its $\Nplus$-action by scaling
endomorphisms, is a semiring object of $\PSh{\Nplus}$: it plays the role
of the ring of functions, recording characteristic-$1$ algebra at every
point.  Their combination is the Arithmetic Site.

\begin{definition}[label=def:arithmetic_site,
                  uses=def:presheaf_topos, def:structure_sheaf]
  \label{def:arithmetic_site}\uses{def:presheaf_topos, def:structure_sheaf}\lean{ArithmeticSite.arithmeticSite}
  The \emph{Arithmetic Site} is the ringed topos $(\PSh{\Nplus}, \mathcal{O})$
  consisting of the presheaf topos $\PSh{\Nplus}$ together with its
  structure sheaf $\mathcal{O} = \Nbar$.
\end{definition}

\begin{remark}
  The Arithmetic Site is a \emph{ringed topos of characteristic $1$}:
  its structure sheaf is a semiring with idempotent addition ($a \oplus
  a = a$), placing the entire construction in the world of tropical
  algebra rather than classical commutative ring theory.  Connes and
  Consani describe it as the ``algebraic geometric incarnation'' of the
  non-commutative geometric approach to the Riemann hypothesis: the site
  provides a geometric framework, analogous to the function-field setting
  of the Weil proof, in which the relevant $L$-function --- the Riemann
  zeta function --- arises as a Hasse--Weil zeta function.  The
  subsequent chapters make this precise by analysing the points of the
  site and the Frobenius correspondences acting on them.
\end{remark}

\chapter{Points of the topos}

In topos theory a \emph{point} of a topos $\mathcal{E}$ generalises the
notion of a point of a topological space: it is a geometric morphism
$p : \Cat{Set} \to \mathcal{E}$, pulling back each sheaf to its stalk.
For a presheaf topos $\widehat{C}$, points correspond to flat functors
$C \to \Cat{Set}$.  For $\PSh{\Nplus} = \widehat{B\Nplus}$ this means
a point is a \emph{flat $\Nplus$-set}: a set $X$ with a $\Nplus$-action
that arises as a filtered colimit of free $\Nplus$-sets.  The remarkable
theorem of Connes--Consani is that such sets are classified, up to
isomorphism, by the ad\`ele class space of $\mathbb{Q}$ --- a
non-commutative space that already appears in Connes' spectral approach
to the Riemann hypothesis.

\begin{definition}[label=def:topos_point, uses=def:presheaf_topos]
  \label{def:topos_point}\uses{def:presheaf_topos}
  A \emph{point} of a topos $\mathcal{E}$ is a geometric morphism
  $p : \Cat{Set} \to \mathcal{E}$, i.e.\ an adjoint pair
  $p^* \dashv p_*$ where $p^*$ is left exact.
\end{definition}

\begin{remark}
  For $\PSh{\Nplus}$, the flat $\Nplus$-sets admit a concrete
  classification.  A $\Nplus$-set $X$ is flat if and only if it is
  isomorphic (as an ordered $\Nplus$-set) to a non-trivial subgroup of
  $(\mathbb{Q}, \mathbb{Q}_+)$ with the action of $n \in \Nplus$ by
  multiplication.  The isomorphism classes of such subgroups are in
  natural bijection with elements of the double quotient
  $\Qplus \backslash \Af / \Zhat^*$: an ad\`ele $a \in \Af$ determines
  a subgroup $H_a = \{ q \in \mathbb{Q} \mid qa \in \Zhat \}$ of
  $\mathbb{Q}$, and two ad\`eles give isomorphic subgroups precisely when
  they lie in the same double coset.  The rational ad\`eles $a \in
  \mathbb{Q}^\times$ yield the single isomorphism class $\mathbb{Z}$,
  which is why the purely rational part of the space corresponds to the
  ``points over $\mathbb{B}$'' in the sense of Connes--Consani.
\end{remark}

\begin{theorem}[label=thm:points_are_adele_classes,
               uses=def:topos_point, def:arithmetic_site]
  \label{thm:points_are_adele_classes}\uses{def:topos_point, def:arithmetic_site}\lean{ArithmeticSite.points_are_adeleClasses}
  The space of isomorphism classes of points of $\PSh{\Nplus}$ is
  canonically isomorphic to the double quotient
  \[
    \Qplus \backslash \Af / \Zhat^*,
  \]
  a piece of the ad\`ele class space of $\mathbb{Q}$.  More precisely,
  the category of points of $\PSh{\Nplus}$ is equivalent to the category
  of totally ordered groups isomorphic to non-trivial subgroups of
  $(\mathbb{Q}, \mathbb{Q}_+)$, with injective order-preserving
  homomorphisms as morphisms.
\end{theorem}
\begin{proof}
  This is Theorem~2.2 of Connes--Consani (2014).  Points of
  $\PSh{\Nplus}$ correspond to flat $\Nplus$-sets, which are identified
  with non-trivial subgroups of $(\mathbb{Q}, \mathbb{Q}_+)$ by a direct
  analysis of the flatness condition.  The double quotient description
  then follows from the identification $H_a = \{ q \in \mathbb{Q} \mid
  qa \in \Zhat\}$ described in the remark above. \sorry
\end{proof}

\chapter{The Frobenius correspondences}
\label{chap:frobenius_correspondences}

The Frobenius correspondences are the analogue, in characteristic $1$,
of the Frobenius endomorphisms of a variety over a finite field.  They
do not act on the Arithmetic Site itself --- that would require
endomorphisms of the ringed topos $(\PSh{\Nplus}, \mathcal{O})$ ---
but rather on its \emph{square} $(\widehat{B{\Nplus}^2},
\Nbar \otimes_{\mathbb{B}} \Nbar)$, where $\mathbb{B} = (\{0,1\},
\max, \times)$ is the Boolean semifield.  The square is obtained by
taking the product topos $\PSh{\Nplus} \times \PSh{\Nplus}$ and
equipping it with the tensor product of structure sheaves.

For each pair $(n, m) \in \Nplus \times \Nplus$, Connes and Consani
define an endomorphism $\mathrm{Fr}_{n,m}$ of $\Nbar \otimes_{\mathbb{B}}
\Nbar$ by
\[
  \mathrm{Fr}_{n,m}\!\left(\sum_i q^{a_i} \otimes q^{b_i}\right)
  := \sum_i q^{na_i} \otimes q^{mb_i},
\]
scaling the two factors independently.  The Frobenius correspondence at
a rational $\lambda = n/m \in \Qplus$ (with $n, m$ coprime) is then
$\Phi_\lambda := \mu \circ \mathrm{Fr}_{n,m}$, where $\mu : \Nbar
\otimes_{\mathbb{B}} \Nbar \to \Nbar$ is the product map
$\mu(q^a \otimes q^b) = q^{a+b}$.  For general $\lambda \in \Qplus$
the correspondence is defined by the formula
\[
  F(\lambda, q)\!\left(\sum_i q^{n_i} \otimes q^{m_i}\right)
  = q^{\,\inf_i(\lambda n_i + m_i)},
\]
and extended to irrational $\lambda$ by diophantine approximation.

\begin{definition}[label=def:frobenius_correspondence,
                  uses=def:arithmetic_site, def:scaling_action]
  \label{def:frobenius_correspondence}\uses{def:arithmetic_site, def:scaling_action}\lean{ArithmeticSite.frobeniusCorrespondence}
  For each $\lambda \in \Qplus$, the \emph{Frobenius correspondence}
  $\Phi_\lambda$ is the reduced correspondence on the square of the
  Arithmetic Site defined by the formula $F(\lambda, q)$ above.  It
  acts on the points of $(\PSh{\Nplus}, \mathcal{O})$ over $[0,1]_{\max}$
  by the scaling map $x \mapsto \lambda x$ on the underlying ad\`ele
  class space.
\end{definition}

\begin{theorem}[label=thm:frobenius_are_correspondences,
               uses=def:frobenius_correspondence,
               thm:points_are_adele_classes]
  \label{thm:frobenius_are_correspondences}\uses{def:frobenius_correspondence, thm:points_are_adele_classes}\lean{ArithmeticSite.frobeniusCorrespondence_comp}
  The Frobenius correspondences compose multiplicatively:
  $\Phi_\lambda \circ \Phi_{\lambda'} = \Phi_{\lambda\lambda'}$ for all
  $\lambda, \lambda' \in \Qplus$ (with a canonical correction term when
  $\lambda, \lambda'$ are both irrational and $\lambda\lambda' \in
  \Qplus$).  In particular, $\lambda \mapsto \Phi_\lambda$ is a
  homomorphism from $(\Qplus, \times)$ into the composition monoid of
  correspondences on the square.
\end{theorem}
\begin{proof}
  This is Theorem~4.3 of Connes--Consani (2014).  The composition law
  follows from the formula for $F(\lambda, q)$ and the identification of
  the range semirings $R(\lambda)$ described in their Section~3.
  \sorry
\end{proof}

\chapter{Connection to the Riemann zeta function}

The Frobenius correspondences $\Phi_\lambda$ act on the points of the
Arithmetic Site over $[0,1]_{\max}$, which are parametrised by the
ad\`ele class space.  Counting the fixed points of $\Phi_\lambda$ as
$\lambda$ varies produces a \emph{counting distribution} $N(u)$, and
the zeta function assembled from $N(u)$ via the Hasse--Weil formula
turns out to be the complete Riemann zeta function.  This is the
central result of Connes--Consani and the main motivation for the
entire construction.

The precise mechanism runs through Connes' distributional trace formula.
If $G = \mathrm{GL}_1(\mathbb{A}_{\mathbb{Q}})/\mathrm{GL}_1(\mathbb{Q})$
is the id\`ele class group and $\vartheta_u\xi(x) = \xi(u^{-1}x)$ is
its scaling action on functions on the ad\`ele class space, then for
test functions $h(u) = g(|u|)$ supported away from $1$ the trace
formula gives
\[
  \mathrm{Tr}_{\mathrm{distr}}\!\left(\int_G h(u)\,\vartheta_u\,d^*u\right)
  = \sum_{v \in \Sigma_{\mathbb{Q}}} \int_{Q_v^*}
    \frac{h(u^{-1})}{|1-u|}\,d^*u,
\]
where the sum runs over all places of $\mathbb{Q}$.  The counting
distribution $N(u)$ is read off from the geometric side of this formula,
and the zeta function is recovered by a standard Mellin-type transform.

\begin{theorem}[label=thm:zeta_counting,
               uses=thm:frobenius_are_correspondences]
  \label{thm:zeta_counting}\uses{thm:frobenius_are_correspondences}\lean{ArithmeticSite.hasseWeilZeta_eq_riemannZeta}
  The zeta function $\zeta_N$ defined by
  \[
    \frac{\partial_s \zeta_N(s)}{\zeta_N(s)}
    = -\int_1^\infty N(u)\,u^{-s}\,d^*u
  \]
  from the counting distribution $N(u)$ of the Frobenius action on
  points of $(\PSh{\Nplus}, \mathcal{O})$ over $[0,1]_{\max}$ is the
  complete Riemann zeta function
  $\zeta_{\mathbb{Q}}(s) = \pi^{-s/2}\,\Gamma(s/2)\,\zeta(s)$.
\end{theorem}
\begin{proof}
  This is Theorem~2.7 of Connes--Consani (2014).  The counting
  distribution $N(u)$ is identified via Theorem~2.6 with the geometric
  term of Connes' trace formula applied to the scaling action of
  $\mathbb{R}^*_+$ on the ad\`ele class space, and the result then
  follows from the computation in Connes--Consani (2010).  \sorry
\end{proof}